Evaluating Russia's Response to the COVID-19 Pandemic: An Analysis of High Excess Death Rates Despite High GHS Index

### 2024-02-15 paper/paper.qmd

---
title: "Evaluating Russia's Preparedness and Response to the COVID-19 Pandemic: An Analysis of Excess Death Rates and the Global Health Security Index"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "This study investigates the unexpected surge in death rates during the COVID-19 pandemic in Russia, a country considered well-prepared according to the Global Health Security (GHS) Index. Building upon Nuzzo and Ledesma's work, we analyze the impact of COVID-19 in Russia, incorporating factors such as the GHS Index and excess death rates. Our findings reveal a substantial increase in excess death rates in Russia, diverging from the lower rates observed in nations deemed well-prepared. This research contributes to our understanding of the repercussions of the COVID-19 outbreak in Russia, highlighting areas for improvement in prepared countries to mitigate the effects of future pandemics."
format: 
  pdf:
    tbl-pos: "h"
    fig-pos: "h"
toc: true
number-sections: true
bibliography: references.bib
---

@@

The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years (World Health Organization [@citeWHOeffect]). With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates (WikiCOVID [@citeWikiCOVID]). The staggering number of deaths has prompted a critical examination of countries presumed to be well-prepared to handle such outbreaks and the factors contributing to their extensive impact.

Among the most severely affected nations, including India, the United States of America, Brazil and Mexico, Russia, a nation deemed well-prepared according to various metrics, was specifically chosen for this study due to its unexpected and disproportionately high death toll in the face of the pandemic. With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia (ThinkGlobalHealth [@citeThinkGlobalHealth]; Carnegie Russia [@citeCarnegieRussia]). This selection aims to uncover unique insights into the challenges faced by ostensibly prepared nations.

This paper replicates the methodology employed by Jennifer B. Nuzzo and Jorge R. Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 (RepPaper [@citeRepPaper]). The choice of GHS is deliberate, considering its comprehensive evaluation of various health security capacities. This metric offers a nuanced understanding of a country's ability to prevent, detect, and respond to public health emergencies, making it a suitable choice over other metrics. However, our focus extends to Russia, providing insights into the impact of the outbreak and proposing improvements for future pandemic mitigation efforts. Despite Russia's apparent preparedness, our reproduction reveals an unprecedented rise in the death rate compared to other countries, with one of the highest excess death rates globally. This prompts a crucial assessment that better planning and regulations are imperative to mitigate the impact of such outbreaks, a key aspect explored in this paper.

The structure of this paper comprises three sections: Data, Results, and Discussion. In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper (RepPaper [@citeRepPaper]). The Results section (@sec-results) summarizes the findings, presenting maps and plots to illustrate trends. The Discussion section (@sec-discussion) delves into the analysis of our findings, exploring measures to prevent extensive effects of future pandemics on the country. Additionally, we discuss limitations and propose avenues for further research, concluding the paper.

@@

## Excess Death Rates {#sec-results-excess-rates}
Considering the likelihood of undercounting COVID-19 deaths, our focus shifted to analyzing excess death rates for the years 2020 and 2021 to provide a comprehensive measure of the pandemic's overall mortality impact. Drawing inspiration from Nuzzo and Ledesma's methodology (@citeRepPaper), which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model. Extrapolating the dashed line of best-fit from 2010-2019 to 2020 and 2021 allowed us to determine the expected death rate for each country. The subsequent plotting of the difference between expected and actual death rates, termed the excess death rate, unveiled insights into the pandemic's impact on mortality.

@@

## Findings: Russia, Excess Deaths and what went wrong during the COVID-19 pandemic {#sec-first-point}

In this paper, we have replicated the paper by Nuzzo and Ledesma [@citeRepPaper], where they focused their analysis on the United States and studying how the world's best prepared country still couldn't control the pandemic. Our paper has instead shifted the focus to another global power and fairly well prepared country Russia, where we found out that Russia, when compared to a subset of other countries:(1) had one of the highest increase in all cause death rate in 2020, and (2) had one of the highest excess death rate caused by the pandemic in 2020 and 2021, as observed in @fig-all-cause-death-rate and @fig-excess-death-rate. Our paper also found and observed the relationship between the GHS index of a country and its excess death rate in @fig-excess-death-rate-ghs, where Russia was seen as an outlier, where it lied a lot higher than the line of best fit implying a higher excess death rate than the other countries around it, even with a similar GHS index.

The main question that this paper tackles is, what exactly did Russia do wrong during the pandemic considering it was the first country to roll out a vaccine? According to a report by Carnegie Endowment for International Peace, by late October 2021, only one-third of the Russian Population had been fully vaccinated against the virus, where other countries like Argentina and China had 56% and 76% of their populations respectively, fully vaccinated by then [@citeCarnegieRussia]. According to reports, one of the main concerns for people being skeptical about the Sputnik V vaccine was that they did not trust that the vaccine would be effective or even safe to use. Another reason for the residents of Russia being skeptical was that Russian President Putin hesitated for months after the release of the vaccine to himself get vaccinated [@citeCarnegieRussia]. I believe that this must have slowed down the rate at which people in Russia were getting vaccinated hence leading to unprecedented deaths as a result.

Another possible reason for Russia not being able to tackle the pandemic well was that the quarantine measures during the outbreak in 2020 and 2021 were generally very sporadic and short [@citeNewYorker]. It was reported that at a time when the cases testing positive reached nearly ten thousand a day, the government put lockdown restrictions but that too only for a week, before everything opened back up. I believe this would have been a huge contributing factor to the extremely high excess deaths as more people would have contracted the virus coming in contact with each, hence spreading it more. Additionally, keeping in mind that at this time, as aforementioned, Russia only had approximately one thirds of its population vaccinated, this would have hit the residents of the country really hard.

In @sec-second-point, we will discuss the measures that Russia could have taken to tackle the pandemic in a better way, and also measures that can be taken for future pandemics.

@@

I believe that, as fast as Russia was at releasing the vaccine, maybe the government could have focused on mass producing and advertising the vaccine as a necessity as well. I believe that as soon as there is any skepticism in regards with vaccines, people will choose not to get one. As mentioned above, this will lead to the virus spreading even faster and eventually to more deaths too. In order to prevent this, the government should show the residents of the country that the vaccine is trustworthy and effective. It is essential to note that in order for this to work, the vaccine's efficacy and safety will also have to be proven to prevent from any ethical implications rising.

It was claimed that at the beginning of the outbreak, Russia was trying to downplay the threat of virus, however, with time they realized how lethal it actually was [@citeCarnegieRussia]. According to me, it is very essential for a country to recognize the potential threat from any epidemic or outbreak and then make laws, and rules and regulations around it to quickly contain the situation before it gets worse. For countries as large as Russia, it might be a particularly challenging task to get such a large population that is spread out under control, but if the right procedures are set in place to tackle such situations, it can be manageable. Additionally, it is essential in these situations to also have firm rules in place for quarantining and lockdowns to prevent the virus from spreading. This is especially even more important in the early stages of the outbreak when there is no vaccine in place to prevent the citizens of a country from contracting the virus. Some countries during the COVID-19 pandemic were very strict with their restrictions and lockdowns, and hence experienced way lesser deaths when compared to other countries which were more relaxed. In the future too, for Russia to be able to control the situation early on, better and more strict regulations will be required.

@@

It is essential to take into account an unintended but highly likely bias that might have taken place in this paper. While working on the paper, we realized that in creating @fig-all-cause-death-rate, @fig-excess-death-rate and @fig-excess-death-rate-ghs, selection bias might have been involved in make decisions on which countries to plot on the graphs. Specifically, we used 4 countries in @fig-all-cause-death-rate and @fig-excess-death-rate to show the unique trends for Russia during the pandemic. Maybe if we chose a set of different countries, some countries would have performed more like Russia and consequently would have impacted our analysis. Similarly, in @fig-excess-death-rate-ghs choosing a different subset of countries might generate a completely different line of best fit which might not have satisfied our exploration in the paper. However, at the same time, it is essential to note that the set of countries chosen in this paper has mostly been random and some countries like India and the United States have been shown constantly to create parallels between the impact of the pandemic on both the countries.

It should be noted that different countries employ varying methods of reporting their data, including mortality rates and excess death rates. Therefore, this study primarily relied on observational data, but it's crucial to account for differential biases in the data. For instance, some countries might underreport their deaths, while others might overreport them. Such discrepancies directly impact the analysis conducted in this paper.

Furthermore, COVID-19 policies were predominantly implemented at the state and local levels, introducing variability throughout the pandemic. Differences in policy implementation among these communities, coupled with the presence of unmeasured differences, complicate the ability to draw causal inferences.

There are a few ethical implications that we found while working on the paper. Mainly, sometimes the data can be incomplete or hidden which is misleading while analyzing that data. It is essential to keep that into account when looking at a report as, if for in this case any data for the deaths were incomplete, that would directly impact how we deem Russia to have tackled the pandemic. This is not just applicable for Russian data, but if other countries' data was missing, it would alter a major part of the analysis done above.

@@

While working on the paper we came across a few weaknesses that can be built on to make the report even more sound. Firstly, we could have tried to take into account the selection bias by choosing the countries except Russia randomly. It is obviously impractical to include all 195 countries in all plots, hence choosing a completely random subset of countries could have reduced the selection bias. However, it is also essential to note that it is preferable to compare Russia with countries with the same infrastructure and stature in the world, as that makes the comparisons more sound. Comparing Russian data with random countries might have led to this paper being less informative and less analytical. Then, another weakness was that while making @fig-map it was observed that the raw data file did not actually contain the GHS indices for all the countries around the world, for example, specifically Greenland was excluded from the dataset, which was the reason why some of the countries in the figure are shaded in grey showing that data was not available for them. If all the data was present for the figure, the analysis could have been even more sound.

As a further scope, an extension to this report could include taking into account the proportion of population vaccinated for a country in 2021. Taking this factor into account, we would be able to clearly point out one of the prominent reasons why Russia might have lacked in their attempt to tackle the pandemic. Alternatively, we can flip the question for this paper around and delve deeper into the countries with the lowest GHS indices which surprisingly did better than expected. This would help us analyze the paper in positive notion, taking into account all the right things they did that led to them tackling the outbreak better than many well-prepared countries.

In conclusion, we have observed that over 2020 and 2021, Russia failed to properly control the spread of the pandemic, and also effectively get its population vaccinated, which consequently lead to a large number of positive cases and deaths around the country. For future epidemics and outbreaks, Russia needs to make strict regulations and procedures which will help them contain the virus and not let it spread in the country and cause enormous amount of deaths.

>>>>

---
title: "Evaluating Russia's Response to the COVID-19 Pandemic: An Analysis of High Excess Death Rates Despite High GHS Index"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "This study investigates the unexpected surge in death rates during the COVID-19 pandemic in Russia, a country considered well-prepared according to the Global Health Security (GHS) Index. Building upon Nuzzo and Ledesma's work, we analyze the impact of COVID-19 in Russia, incorporating factors such as the GHS Index and excess death rates. Our findings reveal a substantial increase in excess death rates in Russia, diverging from the lower rates observed in nations deemed well-prepared. This research contributes to our understanding of the repercussions of the COVID-19 outbreak in Russia, highlighting areas for improvement in prepared countries to mitigate the effects of future pandemics."
format: 
  pdf:
    tbl-pos: "h"
    fig-pos: "h"
toc: true
number-sections: true
bibliography: references.bib
---

@@

The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years [@citeWHOeffect]. With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates [@citeWikiCOVID]. The staggering number of deaths has prompted a critical examination of countries presumed to be well-prepared to handle such outbreaks and the factors contributing to their extensive impact.

Among the most severely affected nations, including India, the United States of America, Brazil and Mexico, Russia, a nation deemed well-prepared according to various metrics, was specifically chosen for this study due to its unexpected and disproportionately high death toll in the face of the pandemic. With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia [@citeThinkGlobalHealth]; [@citeCarnegieRussia]. This selection aims to uncover unique insights into the challenges faced by ostensibly prepared nations.

This paper replicates the methodology employed by Jennifer B. Nuzzo and Jorge R. Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 [@citeRepPaper]. The choice of GHS is deliberate, considering its comprehensive evaluation of various health security capacities. This metric offers a nuanced understanding of a country's ability to prevent, detect, and respond to public health emergencies, making it a suitable choice over other metrics. However, our focus extends to Russia, providing insights into the impact of the outbreak and proposing improvements for future pandemic mitigation efforts. Despite Russia's apparent preparedness, our reproduction reveals an unprecedented rise in the death rate compared to other countries, with one of the highest excess death rates globally. This prompts a crucial assessment that better planning and regulations are imperative to mitigate the impact of such outbreaks, a key aspect explored in this paper.

The structure of this paper comprises three sections: Data, Results, and Discussion. In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper [@citeRepPaper]. The Results section (@sec-results) summarizes the findings, presenting maps and plots to illustrate trends. The Discussion section (@sec-discussion) delves into the analysis of our findings, exploring measures to prevent extensive effects of future pandemics on the country. Additionally, we discuss limitations and propose avenues for further research, concluding the paper.

@@

## Excess Death Rates {#sec-results-excess-rates}
Considering the likelihood of undercounting COVID-19 deaths, our focus shifted to analyzing excess death rates for the years 2020 and 2021 to provide a comprehensive measure of the pandemic's overall mortality impact. Drawing inspiration from Nuzzo and Ledesma's methodology [@citeRepPaper], which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model. Extrapolating the dashed line of best-fit from 2010-2019 to 2020 and 2021 allowed us to determine the expected death rate for each country. The subsequent plotting of the difference between expected and actual death rates, termed the excess death rate, unveiled insights into the pandemic's impact on mortality.

@@

## Findings: Russia, Excess Deaths, and Challenges during the COVID-19 Pandemic {#sec-first-point}

This study, inspired by Nuzzo and Ledesma's work [@citeRepPaper], which scrutinized the United States' pandemic response, redirects the focus towards Russia, a globally influential and reasonably well-prepared nation. Our examination revealed that, in comparison to a subset of other countries, Russia experienced (1) a substantial increase in the all-cause death rate in 2020 and (2) one of the highest excess death rates attributed to the pandemic in 2020 and 2021, as illustrated in @fig-all-cause-death-rate and @fig-excess-death-rate. The exploration of the relationship between a country's Global Health Security (GHS) index and its excess death rate in @fig-excess-death-rate-ghs identified Russia as an outlier, significantly surpassing the line of best fit, indicating a higher excess death rate relative to countries with similar GHS indices.

The primary inquiry arising from this paper pertains to understanding the shortcomings in Russia's pandemic response, especially considering its early deployment of a vaccine. Reports from the Carnegie Endowment for International Peace indicate that, by late October 2021, only one-third of the Russian population had received full vaccination, trailing behind countries like Argentina and China with 56% and 76% vaccination coverage, respectively [@citeCarnegieRussia]. Vaccine skepticism among the Russian populace, fueled by concerns about efficacy and safety, contributed to hesitancy, exacerbated by President Putin's delay in personally receiving the Sputnik V vaccine [@citeCarnegieRussia]. This hesitancy likely impeded the vaccination rate, potentially leading to the unprecedented death toll.

Another plausible factor contributing to Russia's pandemic challenges was the implementation of sporadic and short-lived quarantine measures in 2020 and 2021, as reported by [@citeNewYorker]. Lockdown restrictions, initiated when daily positive cases reached nearly ten thousand, were only enforced for a brief week before being lifted. This lax approach to containment measures may have facilitated widespread transmission, particularly given the low vaccination coverage at that time, amplifying the impact on the population.

In @sec-second-point, we will delve into potential strategies Russia could have employed for a more effective pandemic response and discuss broader measures for addressing future pandemics.

@@

While Russia demonstrated swiftness in developing a vaccine, it could have augmented its efforts in mass production and promotion, emphasizing the vaccine's necessity. Addressing vaccine skepticism is crucial, as public perception significantly influences vaccination rates. To foster trust, the government must transparently demonstrate the vaccine's efficacy and safety, mitigating ethical concerns and bolstering public confidence.

Initial attempts to downplay the virus's severity may have hindered the country's response, as acknowledged later [@citeCarnegieRussia]. Recognizing the gravity of an epidemic is pivotal, prompting proactive legislation and regulations to swiftly contain outbreaks. Despite the logistical challenges posed by Russia's vast population, effective measures and protocols can be established to manage such situations. Robust laws and regulations, coupled with stringent quarantine and lockdown protocols, are imperative, particularly in the absence of a vaccine during the early stages of an outbreak. Countries that implemented strict measures during the COVID-19 pandemic experienced lower mortality rates compared to those with more lenient approaches. For Russia to proactively manage future outbreaks, the establishment of enhanced and stringent regulations is paramount.

@@

Recognizing the potential for unintended biases in our analysis is imperative. In constructing @fig-all-cause-death-rate, @fig-excess-death-rate, and @fig-excess-death-rate-ghs, there is a possibility of selection bias influencing our decision on which countries to include. The choice of countries, including Russia, in these figures might impact the observed trends. A different set of countries could yield alternative patterns and potentially alter our conclusions. However, it is crucial to note that the selection was mostly random, with consistent inclusion of countries like India and the United States to draw meaningful parallels between the pandemic's impact on various nations.

It is essential to acknowledge the inherent variability in how countries report data, including mortality and excess death rates. This study heavily relies on observational data, necessitating an understanding of potential biases. Variations in reporting practices, such as underreporting or overreporting deaths, introduce nuances that directly influence the analysis.

Moreover, the implementation of COVID-19 policies occurred predominantly at the state and local levels, introducing additional complexity and potential unmeasured differences. The diverse approaches among communities make drawing definitive causal inferences challenging.

Ethical considerations also come into play, particularly concerning incomplete or concealed data. Incomplete information can be misleading, impacting the evaluation of a country's pandemic response. This ethical concern extends beyond Russia, encompassing any country with missing or incomplete data, highlighting the need for transparency and accuracy in reporting to ensure robust and unbiased analyses.

@@

During the course of this paper, several weaknesses were identified that could be addressed to enhance its robustness. One notable limitation is the potential for selection bias in country inclusion for analysis. While attempting to mitigate this bias, selecting a random subset of countries excluding Russia could have reduced its impact. However, it's important to note that comparisons with countries of similar infrastructure and global stature provide more meaningful insights. Random selection might have compromised the analytical depth and informativeness of the paper.

Another weakness pertains to the availability of data. In the construction of @fig-map, it was evident that the dataset lacked GHS indices for certain countries, such as Greenland, leading to incomplete shading in the figure. A comprehensive dataset would have strengthened the analysis and provided more nuanced insights into global health security preparedness.

As a future scope, incorporating vaccination rates into the analysis for 2021 could offer valuable insights into pandemic control efforts. Examining the proportion of vaccinated population across countries would help elucidate factors contributing to Russia's challenges in containing the pandemic. Additionally, flipping the focus to countries with lower GHS indices that managed the outbreak effectively could provide lessons in effective pandemic management. Understanding the strategies and interventions that facilitated success in these countries could inform future preparedness and response efforts.

Regarding the selection of countries for comparison, our criteria primarily focused on countries with diverse demographics, varied GHS indices, and significant global influence. While biases may exist, the inclusion of countries like the United States and India aimed to provide a comprehensive analysis of pandemic responses across different contexts. However, future research could explore comparisons with countries of smaller populations and lower GDPs to further enrich the analysis and mitigate potential biases.

In conclusion, the analysis underscores Russia's shortcomings in controlling the COVID-19 pandemic, particularly in vaccination efforts and containment measures. Moving forward, implementing stringent regulations and proactive strategies are imperative to prevent widespread transmission and mitigate the impact of future epidemics and outbreaks.
